\chapter{Exercise 13: Switch Statement}

In other languages like Ruby you have a \ident{switch-statement} that can take
any expression.  Some languages like Python just don't have a
\ident{switch-statement} since an \ident{if-statement} with boolean expressions
is about the same thing.  For these languages, \ident{switch-statements} are
more alternatives to \ident{if-statements} and work the same internally.

The \ident{switch-statement} is actually entirely different and is really "jump
table".  Instead of random boolean expressions, you can only put expressions
that result in integers, and these integers are using to calculate jumps from
the top of the \ident{switch} to the part that matches that value.  Here's some
code that we'll break down to understand this concept of "jump tables":

\begin{code}{ex13.c}
<< d['code/ex13.c|pyg|l'] >>
\end{code}

In this program we take a single command line argument and print out all
of the vowels in an incredibly tedious way to demonstrate
a \ident{switch-statement}.  Here's how the \ident{switch-statement}
works:

\begin{enumerate}
\item The compiler marks the place in the program where the
    \ident{switch-statement} starts, let's call this location Y.
\item It then evaluates the expression in \verb|switch(letter)| to
    come up with a number.  In this case the number will be the
    raw ASCII code of the letter in \verb|argv[1]|.
\item The compiler has also translated each of the \ident{case} 
    blocks like \verb|case 'A':| into a location in the program
    that is that far away.  So the code under \verb|case 'A'| is
    at Y+'A' in the program.
\item It then does the math to figure out where Y+letter is
    located in the \ident{switch-statement}, and if it's too
    far then it adjusts it to Y+default.
\item Once it knows the location, the program "jumps" to that spot
    in the code, and then continues running.  This is why you have
    \ident{break} on some of the \ident{case} blocks, but not others.
\item If \verb|'a'| is entered, then it jumps to \verb|case 'a'|, there's
    no break so it "falls through" to the one right under it \verb|case 'A'|
    which has code and a \ident{break}.
\item Finally it runs this code, hits the break then exits out of the
    \ident{switch-statement} entirely.
\end{enumerate}

This is a deep dive into how the \ident{switch-statement} works, but
in practice you just have to remember a few simple rules:

\begin{enumerate}
\item Always include a \ident{default:} branch so that you catch
    any missing inputs.
\item Don't allow "fall through" unless you really want it, and
    it's a good idea to add a comment \verb|//fallthrough| so 
    people know it's on purpose.
\item Always write the \ident{case} and the \ident{break} before
    you write the code that goes in it.
\item Try to just use \ident{if-statements} instead if you can.
\end{enumerate}

\section{What You Should See}

Here's an example of me playing with this, and also demonstrating
various ways to pass the argument in:

\begin{code}{ex13 output}
\begin{lstlisting}
<< d['code/ex13.out|dexy'] >>
\end{lstlisting}
\end{code}

Remember that there's that \ident{if-statement} at the top that
exits with a \verb|return 1;| when you don't give enough arguments.
Doing a return that's not 0 is how you indicate to the OS that
the program had an error.  Any value that's greater than 0 can be
tested for in scripts and other programs to figure out what
happened.

\section{How To Break It}

It is \emph{incredibly} easy to break a \ident{switch-statement}.
Here's just a few of the ways you can mess one of these up:

\begin{enumerate}
\item Forget a \ident{break} and it'll run two or more
    blocks of code you don't want it to run.
\item Forget a \ident{default} and have it silently
    ignore values you forgot.
\item Accidentally put in variable into the \ident{switch} that
    evaluates to something unexpected, like an \ident{int}
    that becomes weird values.
\item Use initialized values in the test.
\end{enumerate}

You can also break this program in a few other ways.  See if you
can bust it yourself.

\section{Extra Credit}

\begin{enumerate}
\item Write another program that uses math on the letter to
    convert it to lowercase, and then remove all the extraneous
    uppercase letters in the switch.
\item Use the \verb|','| (comma) to initialize \ident{letter} 
    in the \ident{for-loop}.
\item Make it handle all of the arguments you pass it with 
    yet another \ident{for-loop}.
\item Convert this \ident{switch-statement} to an \ident{if-statement}.
    Which do you like better?
\end{enumerate}

