\chapter{Exercise 26: Write A First Real Program}

You are at the half-way mark in the book, so you need to take a 
mid-term.  In this mid-term you're going to recreate a piece
of software I wrote specifically for this book called \program{devpkg}.
You'll then extend it in a few key ways and improve the code, most
importantly by writing some unit tests for it.

\begin{aside}{WARNING: Beta Draft Content}
I wrote this exercise before writing some of the exercises you might
need to complete this.  If you are attempting this one now, please
keep in mind that the software may have bugs, that you might have
problems because of my mistakes, and that you might not know everything
you need to finish it.  If so, tell me at help@learncodethehardway.org
and then wait until I finish the other exercises.
\end{aside}

\section{What Is \program{devpkg}?}

\program{Devpkg} is a simple C program that installs other software.  I made it
specifically for this book as a way to teach you how a real software project is
structured, and also how to reuse other people's libraries.  It uses a
portability library called \href{http://apr.apache.org/}{The Apache Portable
Runtime (APR)} that has many handy C functions which work on tons of platforms,
including Windows.  Other than that, it just grabs code from the internet
(or local files) and does the usual \verb|./configure ; make ; make install|
every programmer does.

Your goal in this exercise is to build \program{devpkg} from source, 
finish each \emph{Challenge} I give, and use the source to understand
what \program{devpkg} does and why.


\subsection{What We Want To Make}

We want a tool that has three commands:

\begin{enumerate}
\item[devpkg -S] Sets up a new install on a computer.
\item[devpkg -I] Installs a piece of software from a URL.
\item[devpkg -L] Lists all the software that's been installed.
\item[devpkg -F] Fetches some source code for manual building.
\item[devpkg -B] Builds fetches source code and installs it, even if already installed.
\end{enumerate}

We want \program{devpkg} to be able to take almost any URL, figure
out what kind of project it is, download it, install it, and register
that it downloaded that software.  We'd also like it to process a
simple dependency list so it can install all the software that a
project might need as well.

\subsection{The Design}

To accomplish this goal \program{devpkg} will have a very simple design:

\begin{description}
\item[Use external commands] You'll do most of the work through external
    commands like \program{curl}, \program{git}, and \program{tar}.  This
    reduces the amount of code \program{devpkg} needs to get things done.
\item[Simple File Database] You could easily make it more complex, but to
    start you'll just make a single simple file database at
    \file{/usr/local/.devpkg/db} to keep track of what's installed.
\item[/usr/local Always] Again you could make this more advanced, but for
    starters just assume it's always \file{/usr/local} which is a standard
    install path for most software on Unix.
\item[configure, make, make install] It's assumed that most software can
    install with just a \program{configure; make; make install} and maybe
    \program{configure} is optional.  If the software can't at a minimum
    do that, then there's some options ot modify the commands, but otherwise
    \program{devpkg} won't bother.
\item[The User Can Be root] We'll assume the user can become root using
    sudo, but that they don't want to become root until the end.
\end{description}

This will keep our program small at first and work well enough to get it
going, at which point you'll be able to modify it further for this exercise.


\subsection{The Apache Portable Runtime}

One more thing you'll do is leverage the \href{http://apr.apache.org/}{The
Apache Portable Runtime (APR)} libraries to get a good set of portable routines
for doing this kind of work.  The APR isn't necessary, and you could probably
write this program without them, but it'd take more code than necessary.
I'm also forcing you to use APR now so you get used to linking and using
other libraries.  Finally, the APR also works on \emph{Windows} so your
skills with it are transferable to many other platforms.

You should go get both the \library{apr-1.4.5} and the \library{apr-util-1.3}
libraries, as well as browse through the documentation available at the 
\href{http://apr.apache.org/}{main APR site at apr.apache.org}.

Here's a shell script that will install all the stuff you need.  You
should write this into a file by hand, and then run it until it can
install APR without any errors.

\begin{code}{APR Install Script}
<< d['code/ex26.1.sh|pyg|l'] >>
\end{code}

I'm having you write this script out because this is basically what
we want \program{devpkg} to do, but with extra options and checks.
In fact, you could just do it all in shell with less code, but then
that wouldn't be a very good program for a C book would it?

Simply run this script and fix it until it works, then you'll have the
libraries you need to complete the rest of this project.


\section{Project Layout}

You need to setup some simple project files to get started.  Here's how I
usually craft a new project:

\begin{code}{Project Skeleton Directory}
<< d['code/ex26.2.sh|pyg|l'] >>
\end{code}

\subsection{Other Dependencies}

You should have already installed APR and APR-util, so now you need 
a few more files as basic dependencies:

\begin{enumerate}
\item \file{dbg.h} from Exercise 20.
\item \file{bstrlib.h} and \file{bstrlib.c} from \href{http://bstring.sourceforge.net/}{http://bstring.sourceforge.net/}. Download the .zip file, extract it, and copy just those two files out.
\item Type \verb|make bstrlib.o| and if it doesn't work, read the "Fixing bstring" 
    instructions below.
\end{enumerate}

\begin{aside}{Fixing bstring}
In some platforms the bstring.c file will have an error like:

\begin{lstlisting}
bstrlib.c:2762: error: expected declaration specifiers or '...' before numeric constant
\end{lstlisting}

This is from a bad define the authors added which doesn't work always.  Just
delete this \verb|#ifdef| then recompile and it should work.
\end{aside}

When that's all done, you should have a \file{Makefile}, \file{README},
\file{dbg.h}, \file{bstrlib.h}, and \file{bstrlib.c} ready to go.

\section{The Makefile}

A good place to start is the \file{Makefile} since this lays out 
how things are built and what source files you'll be creating.

\begin{code}{Makefile}
\begin{lstlisting}
<< d['code/ex26/Makefile|dexy'] >>
\end{lstlisting}
\end{code}

There's nothing in this that you haven't seen before, except maybe the strange
\verb|?=| syntax, which says "set PREFIX equal to this unless PREFIX is already
set".

\section{The Source Files}

From the make file, we see that there's four dependencies for \program{devpkg}
which are:

\begin{description}
\item[bstrlib.o] Comes from \file{bstlib.c} and header file \file{bstlib.h} which
    you already have.
\item[db.o] From \file{db.c} and header file \file{db.h}, and it
    will contain code for our little "database" routines.
\item[shell.o] From \file{shell.c} and header \file{shell.h}, with a couple
    functions that make running other commands like \program{curl} easier.
\item[commands.o] From \file{command.c} and header \file{command.h}, and
    contains all the commands that \program{devpkg} needs to be useful.
\item[devpkg] It's not explicitly mentioned, but instead is the target
    (on the left) in this part of the \file{Makefile}. It comes from
    \file{devpkg.c} which contains the \func{main} function for the whole
    program.
\end{description}

Your job is to now create each of these files and type in their code
and get them correct.

\begin{aside}{Don't Be Fooled By The Magic Show}
You may read this description and think, "Man! How is it that Zed is
so smart he just sat down and typed these files out like this!? I
could never do that."  I didn't magically craft \program{devpkg} 
in this form with my awesome code powers.  Instead, what I did is this:

\begin{enumerate}
\item I wrote a quick little README to get an idea of how I wanted it
    to work.
\item I created a simple bash script (like the one you did) to figure
    out all the pieces that you need.
\item I made one .c file and hacked on it for a few days working through
    the idea and figuring it out.
\item I got it mostly working and debugged, \emph{then} I started
    breaking up the one big file into these four files.
\item After getting these files laid down, I renamed and refined the
    functions and data structures so they'd be more logical and "pretty".
\item Finally, after I had it working the \emph{exact same} but with
    the new structure, I added a few features like the \program{-F} and
    \program{-B} options.
\end{enumerate}

You're reading this in the order I want to teach it to you, but don't think
this is how I always build software.  Sometimes I already know the subject and
I use more planning.  Sometimes I just hack up an idea and see how well it'd
work.  Sometimes I write one, then throw it away and plan out a better one.  It
all depends on what my experience tells me is best, or where my inspiration
takes me.

If you run into an "expert" who tries to tell you there's only one
way to solve a programming problem, then they're lying to you.  Either
they actually use multiple tactics, or they're not very good.
\end{aside}

\subsection{The DB Functions}

There must be a way to record URLs that have been installed, list these
URLs, and check if something has already been installed so we can
skip it.  I'll use a simple flat file database and the \file{bstrlib.h}
library to do it.

First, create the \file{db.h} header file so you know what you'll be
implementing.

\begin{code}{db.h}
<< d['code/ex26/db.h|pyg|l'] >>
\end{code}

Then implement those functions in \file{db.c}, as you build this, use
\program{make} like you've been to get it to compile cleanly.

\begin{code}{db.c}
<< d['code/ex26/db.c|pyg|l'] >>
\end{code}

\subsubsection{Challenge 1: Code Review}

Before continuing, read every line of these files carefully and 
confirm that you have them entered in \emph{exactly}.  Read them
line-by-line backwards to practice that. Also trace each function 
call and make sure you are using \func{check} to validate the
return codes.  Finally, look up \emph{every} function that you
don't recognize either on the APR web site documentation, or
in the \file{bstrlib.h} and \file{bstrlib.c} source.


\subsection{The Shell Functions}

A key design decision for \program{devpkg} is to do most of the work
using external tools like \program{curl}, \program{tar}, and \program{git}.
We could find libraries to do all of this internally, but it's pointless
if we just need the base features of these programs.  There is no shame
in running another command in Unix.

To do this I'm going to use the \file{apr\_thread\_proc.h} functions
to run programs, but I also want to make a simple kind of "template"
system.  I'll use a \ident{struct Shell} that holds all the information
needed to run a program, but has "holes" in the arguments list where I
can replace them with values.

Look at the \file{shell.h} file to see the structure and the commands I'll use.
You can see I'm using \ident{extern} to indicate that other \file{.c} files
can access variables I'm defining in \file{shell.c}.

\begin{code}{shell.h}
<< d['code/ex26/shell.h|pyg|l'] >>
\end{code}

Make sure you've created \file{shell.h} exactly, and that you've got the
same names and number of \ident{extern Shell} variables.  Those are used
by the \func{Shell\_run} and \func{Shell\_exec} functions to run commands.
I define these two functions, and create the real variables in \file{shell.c}.

\begin{code}{shell.c}
<< d['code/ex26/shell.c|pyg|l'] >>
\end{code}

Read the \file{shell.c} from the bottom to the top (which is a common C source
layout) and you see I've created the actual \ident{Shell} variables that you
indicated were \ident{extern} in \file{shell.h}.  They live here, but are 
available to the rest of the program.  This is how you make global variables
that live in one \file{.o} file but are used everywhere.  You shouldn't make
many of these, but they are handy for things like this.

Continuing up the file we get to the \func{Shell\_run} function, which is
a "base" function that just runs a command based on what's in a \ident{Shell}
struct.  It uses many of the functions defined in \file{apr\_thread\_proc.h}
so go look up each one to see how it works.  This seems like a lot of work
compared to just using the \func{system} function call, but this also gives
you more control over the other program's execution.  For example, in our
\ident{Shell} struct we have a \ident{.dir} attribute which forces the program
to be in a specific directory before running.

Finally, I have the \func{Shell\_exec} function, which is a "variable arguments"
function.  You've seen this before, but make sure you grasp the \file{stdarg.h}
functions and how to write one of these.  In the challenge for this section
you are going to analyze this function.

\subsubsection{Challenge 2: Analyze Shell\_exec}

Challenge for these files (in addition to a full code review just like you
did in Challenge 1) is to fully analyze \func{Shell\_exec} and break down
exactly how it works.  You should be able to understand each line, how
the two \ident{for-loops} work, and how arguments are being replaced.

Once you have it analyzed, add a field to \ident{struct Shell} that gives
the number of variable \ident{args} that must be replaced.  Update all the
commands to have the right count of args, and then have an error check that
confirms these args have been replaced and error exit.

\subsection{The Command Functions}

Now you get to make the actual commands that do the work.  These commands
will use functions from APR, \file{db.h} and \file{shell.h} to do the 
real work of downloading and building software you want it to build.
This is the most complex set of files, so do them carefully.  As before, you
start by making the \file{commands.h} file, then implementing its functions
in the \file{commands.c} file.

\begin{code}{commands.h}
<< d['code/ex26/commands.h|pyg|l'] >>
\end{code}

There's not much in \file{commands.h} that you haven't seen already.  You
should see that there's some defines for strings that are used everywhere.
The real interesting code is in \file{commands.c}.

\begin{code}{commands.c}
<< d['code/ex26/commands.c|pyg|l'] >>
\end{code}

After you have this entered in and compiling, you can analyze it.  If you've
don the challenges until now, you should see how the \file{shell.c} functions
are being used to run shells and how the arguments are being replaced.  If
not then go back and make sure you \emph{truly} understand how \func{Shell\_exec}
actually works.

\subsubsection{Challenge 3: Critique My Design}

As before, do a complete review of this code and make sure it's exactly
the same.  Then go through each function and make sure you know how it 
works and what it's doing.  You also should trace how each function calls
the other functions you've written in this file and other files.  Finally,
confirm that you understand all the functions you're calling from APR here.

Once you have the file correct and analyzed, go back through and assume
I'm an idiot.  Then, criticize the design I have to see how you can improve
it if you can.  Don't \emph{actually} change the code, just create a little
\file{notes.txt} file and write down your thoughts and what you might change.


\subsection{The \file{devpkg} Main Function}

The last and most important file, but probably the simplest, is \file{devpkg.c}
where the \func{main} function lives.  There's no \file{.h} file for this, since
this one includes all the others.  Instead this just creates the executable
\program{devpkg} when combined with the other \file{.o} files from our 
\file{Makefile}.  Enter in the code for this file, and make sure it's 
correct.

\begin{code}{devpkg.c}
<< d['code/ex26/devpkg.c|pyg|l'] >>
\end{code}

\subsubsection{Challenge 4: The README And Test Files}

The challenge for this file is to understand how the arguments are
being processed, what the arguments are, and then create the \file{README}
file with instructions on how to use it.  As you write the README, also
write a simple \file{test.sh} that runs \program{./devpkg} to check that
each command is actually working against real live code.  Use the \verb|set -e|
at the top of your script so that it aborts on the first error.

Finally, run the program under valgrind and make sure it's all working
before moving on to the mid-term exam.

\section{The Mid-Term Exam}

Your final challenge is the mid-term exam and it involves three things:

\begin{enumerate}
\item Compare your code to my code available online and starting with 100\%, 
    remove 1\% for each line you got wrong.
\item Take your notes.txt on how you would improve the code and functionality
    of \program{devpkg} and implement your improvements.
\item Write an alternative version of \program{devpkg} using your other 
    favorite language or the one you think can do this the best.  Compare
    the two, then improve your \emph{C} version of \program{devpkg} based on what
    you've learned.
\end{enumerate}

To compare your code with mine, do the following:

\begin{lstlisting}
cd ..  # get one directory above your current one
git clone git://gitorious.org/devpkg/devpkg.git devpkgzed
diff -r devpkg devpkgzed
\end{lstlisting}

This will clone my version of \program{devpkg} into a directory
\program{devpkgzed} and then use the tool \program{diff} to compare
what you've done to what I did.  The files you're working with in
this book come directly from this project, so if you get different
lines then that's an error.

Keep in mind that there's no real pass or fail on this exercise, just
a way for you to challenge yourself to be as exact and meticulous as
possible.

